---
title: 'Task 1: Spatial Data Visualization'
description: Spatial Visualization of California Oil Spills in 2008
author: 
  - name: Shayan Kaveh
date: 2022-03-03
output: 
  distill::distill_article:
    theme: yeti
    toc: yes
    toc_depth: 2
    toc_float: yes
    collapsed: yes
    smooth_scroll: yes
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include = T, message = F, warning = F}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
library(tidyverse)
library(here)
library(broom)
library(sf)
library(tmap)
library(plotly)
```

## Overview

This analysis visualizes data from the Oil Spill Prevention and Response (OSPR) Incident Tracking database, collected by OSPR Field Response team members. The analysis looks at overall oil spills and a closer look at inland oil spills in 2008. 

**Data Source:** Lampinen, Mark (2009). Oil Spill Incident Tracking [ds394]. California Department of Fish and Game, Office of Spill Prevention and Response. https://map.dfg.ca.gov/metadata/ds0394.html?5.108.39

## Analysis

```{r data}
ca_counties_sf <- read_sf(here("_posts", "2022-03-05-spatial-data", "data", "ca_counties", "CA_Counties_TIGER2016.shp")) %>% 
  janitor::clean_names() %>% 
  select(county_name = name)

# st_crs(ca_counties_sf) ### WGS84, epsg: 3857

ca_oil_sf <- read_sf(here("_posts", "2022-03-05-spatial-data", "data", "ds394", "ds394.shp")) %>% 
  janitor::clean_names()

# st_crs(ca_oil_sf) ### NAD83, epsg: 3310

oil_3857_sf <- st_transform(ca_oil_sf, st_crs(ca_counties_sf))

# st_crs(oil_3857_sf) 
#wooh it's in wgs 84 now
```

### All oil spills (2008)
```{r tmap, fig.align = "center"}
tmap_mode(mode = "view")

tm_shape(ca_counties_sf) +
  tm_fill() +
  tm_basemap(c(StreetMap = "OpenStreetMap",
               TopoMap = "OpenTopoMap")) +
  tm_borders() +
  tm_polygons(alpha = 0) + # remove county fill
  tm_shape(ca_oil_sf) +
  tm_dots(col = "indianred")
```
Figure 1. Interactive California map representing all oil spills in 2008

<br> 

### Oil spills by county
```{r spatial data}
oil_ca_sf <- ca_counties_sf %>% 
  st_join(oil_3857_sf)
```

```{r}
oil_counts_sf <- oil_ca_sf %>% 
  group_by(county_name) %>% 
  filter(inlandmari %in% "Inland") %>% 
  summarize(sum_spills = sum(!is.na(oesnumber))) # summarize total oil spills by county

# head(oil_counts_sf)
```

```{r, fig.align = "center", fig.cap = "Figure 2. Total number of inland oil spills summarized by county"}
chloropleth <- ggplot(data = oil_counts_sf) +
  geom_sf(aes(fill = sum_spills), color = "gray", size = 0.5) +
  scale_fill_gradientn(colors = c("ivory", "lightcoral", "indianred4")) +
  labs(fill = "Number of inland oil spills by county") +
  theme_void() 

ggplotly(chloropleth, tooltip = c("County", "sum_spills"))
```

